---
title: "Past hour data"
output: html_document
date: "2026-08-04"
---


```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```


## Exploration

### Overview
```{r}
# Read raw dataset
hourly_raw <- read_csv(
  "../data/raw/pedestrian-counting-system-monthly-counts-per-hour.csv",
  show_col_types = FALSE
) |>
  clean_names()
```

```{r}
# overview
dim(hourly_raw)
names(hourly_raw)
glimpse(hourly_raw)
head(hourly_raw)
```


```{r}
# check the whole during
hourly_raw %>%
  summarise(
    earliest = min(sensing_date),
    latest = max(sensing_date)
  )
```
```{r}
# count records by year
hourly_raw |>
  mutate(
    year = year(sensing_date)
  ) |>
  count(year)
```
#### Observation:
The dataset covers pedestrian counts from 2024 to 2026. Most records were collected during 2025, while 2024 and 2026 contain fewer records because they represent partial years.

```{r}
# count unique sensors
hourly_raw |>
  summarise(
    unique_sensors = n_distinct(location_id)
  )
```

#### Observation:
The dataset contains data from 103 unique pedestrian sensors, indicating that hourly pedestrian counts were collected across multiple monitoring locations within Melbourne.


```{r}
# overview of the pedestrian count
summary(hourly_raw$total_of_directions)
```
#### Observation:
Hourly pedestrian counts range from 0 to 11,113, with a median of 161 and a mean of 393.9. The mean is substantially higher than the median, indicating a right-skewed distribution where a small number of observations have exceptionally high pedestrian counts.


#### Dataset Overview:
The Pedestrian Counting System (Counts per Hour) dataset contains 1,613,233 records and 9 attributes, covering the period from 4 August 2024 to 3 August 2026. Each record represents the pedestrian count recorded at a specific sensor location during a particular hour. The dataset will be used as the historical pedestrian count dataset for the project.


## Missing checking
```{r}
# check missing data
colSums(is.na(hourly_raw))
```
#### Observation:
No missing values were found in the key attributes (location_id, sensing_date, hour_day, direction_1, direction_2, and total_of_directions). Missing values only occur in sensor_name and location, with 25,593 records each. These attributes are not essential for the MVP because sensor descriptions and coordinates are already available in the cleaned Sensor Locations dataset.


```{r}
# check duplicate records
hourly_raw |>
  count(location_id, sensing_date, hour_day) |>
  filter(n > 1)
```
#### Observation:
No duplicate records were found based on the combination of location_id, sensing_date, and hour_day. This confirms that each sensor has only one hourly pedestrian count record for a given date and hour.


## Exploration Summary:
The dataset contains 1,613,233 hourly pedestrian count records collected from 103 pedestrian sensors between August 2024 and August 2026. No missing values or duplicate records were found in the key attributes required for analysis. Missing values are limited to sensor_name and location, which are not required for the MVP. Overall, the dataset is complete and suitable for further data cleaning and analysis.



## Data Wrangling
### Data Wrangling Strategy
To support the MVP, the hourly pedestrian count dataset will be transformed into a historical baseline rather than retaining all raw records. First, the dataset will be filtered to a complete one-year period (August 2025 – July 2026). Temporal features, including weekday, weekend, and public holiday, will then be derived from the date. Finally, the data will be grouped by sensor location, day type, and hour to calculate the average pedestrian count. This processed dataset will provide a meaningful historical reference for the application's real-time pedestrian density visualisation and route recommendation features.


#### Step 1: Filter a complete one-year period
To establish a consistent historical baseline, the dataset was restricted to a complete 12-month period from 1 August 2025 to 31 July 2026. This avoids using partial years while preserving seasonal variations in pedestrian activity.

```{r}
# Select the time period
hourly_1year <- hourly_raw |>
  filter(
    sensing_date >= as.Date("2025-08-01"),
    sensing_date <= as.Date("2026-07-31")
  )
```

```{r}
# count the records in one year
hourly_1year |>
  summarise(
    earliest = min(sensing_date),
    latest = max(sensing_date),
    records = n()
  )
```

#### Step 2：Create new time attributes(Weekday, weekday_num)
```{r}
hourly_time <- hourly_1year |>
  mutate(
    # Create weekday number: Monday = 1, Sunday = 7
    weekday_num = wday(
      sensing_date,
      week_start = 1
    ),
    # Create weekday name in English
    weekday = wday(
      sensing_date,
      label = TRUE,
      abbr = FALSE,
      week_start = 1,
      locale = "English"
    )
  )
```


```{r}
# Check the new time attributes
hourly_time |>
  select(sensing_date, weekday_num, weekday) |>
  head(5)
```

```{r}
# count records by weekday
hourly_time |>
  count(weekday)
```

#### Observation:
The weekend indicator was successfully created. Of the 825,104 records, 588,525 were recorded on weekdays and 236,579 on weekends.


#### Step3 Select required columns
```{r}
hourly_selected <- hourly_time |>
  select(
    location_id,
    sensing_date,
    weekday,
    weekday_num,
    hour_day,
    direction_1,
    direction_2,
    total_of_directions
  )
head(hourly_selected, 5)
```


#### The dataset was reduced to the attributes required for the MVP.

#### Kept

location_id (join key)
sensing_date (date)
weekday (derived temporal feature)
is_weekend (derived temporal feature)
hour_day (hour of day)
direction_1 (pedestrian flow direction)
direction_2 (pedestrian flow direction)
total_of_directions (hourly pedestrian count)

#### Dropped

id (unique record identifier, not required)
sensor_name (available in the Sensor Locations dataset)
location (coordinates provided in the Sensor Locations dataset)

#### Step 4. Create Historical Baseline

```{r}
hourly_baseline <- hourly_selected |>
  group_by(
    location_id,
    weekday_num,
    weekday,
    hour_day
  ) |>
  summarise(
    # Overall pedestrian counts
    average_count = round(
      mean(total_of_directions, na.rm = TRUE),
      2
    ),

    median_count = median(
      total_of_directions,
      na.rm = TRUE
    ),

    # Direction 1 baseline
    d1_med = median(
      direction_1,
      na.rm = TRUE
    ),

    # Direction 2 baseline
    d2_med = median(
      direction_2,
      na.rm = TRUE
    ),

    # Overall per-minute baseline
    total_med_perminute = round(
      median(total_of_directions, na.rm = TRUE) / 60,
      2
    ),

    # Directional per-minute baseline
    d1_med_perminute = round(
      median(direction_1, na.rm = TRUE) / 60,
      2
    ),

    d2_med_perminute = round(
      median(direction_2, na.rm = TRUE) / 60,
      2
    ),

    sample_size = n(),

    .groups = "drop"
  ) |>
  arrange(
    location_id,
    weekday_num,
    hour_day
  )
head(hourly_baseline,5)
```

#### For each group, the median pedestrian count was calculated to represent the typical pedestrian activity. The median was selected instead of the mean because it is less affected by abnormal events (e.g., festivals or major public events) and provides a more stable historical reference.

#### In addition, the number of observations (sample_size) was retained to indicate how many historical records contributed to each baseline value.


```{r}
dim(hourly_baseline)

```
#### Observation

The aggregation reduced the dataset from 825,104 hourly records to 17,037 historical baseline records, providing a lightweight dataset suitable for real-time comparison and route recommendation.


```{r}
# Summarise pedestrian counts by weekday
weekday_raw_summary <- hourly_selected |>
  group_by(weekday) |>
  summarise(
    # Number of records for each weekday
    record_count = n(),
    
    # Minimum hourly pedestrian count
    minimum_count = min(total_of_directions, na.rm = TRUE),
    
    # Maximum hourly pedestrian count
    maximum_count = max(total_of_directions, na.rm = TRUE),
    
    # Median hourly pedestrian count
    median_count = median(total_of_directions, na.rm = TRUE),
    .groups = "drop"
  )

weekday_raw_summary
```


```{r}
summary(hourly_baseline$sample_size)
```

```{r}
sum(is.na(hourly_baseline))
```

### Validation
```{r}
hourly_baseline |>
  count(location_id) |>
  summary()
```

#### Observation:
Most sensor locations contain the expected 168 historical baseline records (7 weekdays × 24 hours). A small number of sensors have fewer records, likely due to missing observations or periods when the sensors were inactive. Overall, the dataset provides comprehensive temporal coverage for historical baseline construction.

#### Check the minimum record sensor
```{r}
# find Min: 1.00
hourly_baseline |>
  filter(sample_size == 1)
```


```{r}
# check the NO.37 sensor
hourly_selected |>
  filter(location_id == 37) |>
  summarise(
    earliest = min(sensing_date),
    latest = max(sensing_date),
    records = n()
  )
```
#### Observation:
Sensor 37 covers the full one-year period but contains only 5,965 hourly records, compared with a theoretical maximum of 8,760. This indicates that the sensor has substantial gaps in hourly observations. The records were retained, while sample_size was preserved to indicate the amount of historical data supporting each median baseline.


```{r}
# check duplication
hourly_baseline |>
  count(location_id, weekday, hour_day) |>
  filter(n > 1)
```
No duplicate

### Export
```{r}
write_csv(
  hourly_baseline,
  "../data/processed/historical_crowd_baseline.csv"
)
```



### Review
```{r}
baseline_check <- read_csv(
  "../data/processed/historical_crowd_baseline.csv",
  show_col_types = FALSE
)

glimpse(baseline_check)

head(baseline_check)
```

#### Validation

The processed dataset was validated before export.

The following checks were performed:

No missing values.
No duplicated combinations of location_id, weekday, and hour_day.
Sample sizes were examined to ensure that most baseline values were calculated from approximately one year of historical observations.
The processed dataset was successfully exported and reloaded for verification.








